### 2022-06-10 paper.tex
going through Leonies part

Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells. For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citet{bird1986cpg}. In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data Figure \ref{workflow}. The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}. For each cell, these tools produce a text file that lists the methylation status of all CpG sites. Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.

\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e. the number of CpG sites that have sequencing coverage. These summary statistics can be used to detect cells with poor quality. The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS). The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes. Cells that do not show this pattern, or cells with few observed CpG sites, may then be filtered from the data set with \texttt{scbs filter}.

After quality control, the user has access to the previously described genome-wide VMR detection approach via \texttt{scbs scan}. This produces a BED-file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to an scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\section{Comparison}

To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
First, we use a data set by \citet{argelaguet2019gastru}. In their work, they generated a single cell methylome data set of mouse embryo cells across different stages of development (E4.5 – E7.5). They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development. As promoters, they used DNA regions of \textpm2~kb around transcriptions start sites (TSS).

--The following paragraph potentially needs to be changed when new plots are made--

>>>>

Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells. For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}. In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{workflow}). The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}. For each cell, these tools produce a text file that lists the methylation status of all CpG sites. Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.

In brief, we store methylation of each chromosome in a binary matrix where each column represents a cell and each row represents a base pair. Since the vast majority of values in this matrix are missing due to the sparsity of scBS data, we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format. This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e. the number of CpG sites that have sequencing coverage. These summary statistics can be used to detect cells with poor quality. The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution. The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes. Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{scbs filter}.

After quality control, the user has access to the previously described genome-wide VMR detection approach via \texttt{scbs scan}. This produces a BED-file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\section{Comparison}
To assess performance of our newly devised methods, we obtained two previously published scBS data sets and re-analysed them with \texttt{scbs}.

The first data set is a single-cell multi-omic data set generated by \citet{argelaguet2019gastru}. In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5 – E7.5), as well as single-cell transcriptomes from the same cells. They calculated the correlation between RNA expression and DNA methylation at the corresponding promoter region to elucidate the implications of the epigenome on gene expression in development. \citet{argelaguet2019gastru} defined promoters as genomic intervals \textpm2~kb around the TSS of each gene.
